Simple repression draft

### 2020-05-10 doc/paper.tex
Small edits + outline sketch of section on means & fold-change

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As shown in Figure~1.
\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.pdf}
\caption{\textbf{
    An overview of the simple repression motif at the level of means.}
    (A) depicts the qualitative biological picture of simple repression.
    (B) shows the ``master curve'' to which all cartoons in (C) and (D) collapse.
    (C) shows two possible equilibrium pictures of simple repression, while
    (D) shows four possible coarse-graining in a nonequilibrium picture.
    }
\label{fig1:means_cartoons}
\end{figure}

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
\caption{\textbf{Constitutive promoter cartoon comparison.}
The left column depicts various plausible cartoons for constitutive promoters,
along with literature references to their prior usage.
In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
Model (2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}.
Model (3) is a more detailed generalization of model (2),
treating transcription initiation as a multi-step process
proceeding through closed and open complexes~\cite{Mitarai2015}.
(4) a more coarse-grained version of (2) where the precise nature
of active and inactive states are left ambiguous~\mmnote{Cite Manuel}, and
(5) transcripts are produced in bursts,
and initiation of bursts is a Poisson process.
\mmnote{Haven't found model 5 formulated quite like this in the literature,
usually burstiness is handled like model 4, but I should probably search a bit more.}
The right column shows the Fano factor for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$
strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data
observed in~\cite{Jones2014} without introducing additional
assumptions and model complexity.
}
\label{fig:constit_cartoons}
\end{figure}

@@

In fact, we suspect \textit{any} model in which transcription
proceeds through a multistep cycle must necessarily have $\nu<1$.
The intuitive argument compares the waiting time distribution to
traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple
exponentials, and intuitively should be more peaked with a
smaller fractional width than an exponential with the same mean.\footnote{
This can be made more precise with a result
from~\cite{Stewart2007}, who showed that the convolution of
multiple gamma distributions (of which the exponential
distribution is a special case) is to a very good approximation
also gamma distributed. Using their Eq.~2 for the distribution of
the convolution, with shape parameters set to 1 to
give exponential distributions, the total waiting time
distribution has variance to squared mean ratio
$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
where the $k_i$ are the rates of the individual steps. Clearly
this is less than 1 and therefore the total waiting time
distribution is narrower than the corresponding exponential.} A
less disperse waiting time distribution means transcription
initations are more uniformly distributed in time relative to a
Poisson process. Hence the distribution of mRNA over a population
of cells should be less variable compared to Poisson, giving
$\nu<1$.\footnote{
This last step, while intuitive, can be argued by analogy to
photon statistics where antibunching gives rise to sub-Poissonian
noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
not expect they apply for our problem. Nevertheless we refrain
from elevating to a ``theorem'' this cycle/sub-Poissonian equivalence.
\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
be totally out of left field for most readers, any suggestions on
a more comprehensible reference??}
}

@@

While this seems promising, there is a major drawback for our purposes
which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
the statistical inference problem is nonidentifiable
\mmnote{find a good cite for this usage},
meaning it is impossible to infer the
parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
Rather, only the ratio $r/k^-$ could be inferred. In that work,
the problem was worked around with an informative prior on the
ratio $k^-/k^+$. That approach is unlikely to work here, as,
recall, our entire goal in modeling constitutive expression is to
use it as the basis for a yet more complicated model, when we add
on repression. But adding more complexity to a model that is
already poorly identified is a fool's errand,
so we will explore one more potential model.

>>>>

\mmnote{Key ideas, no particular order, that I haven't written down before:
\begin{itemize}
\item Our goal is to build phenomenological models of input-output
functions of genetic circuits. We want concepts, not mere facts.
In particular a key question is: at this level of
coarse graining, what microscopic details do I need to explicitly
model, and how do we figure that out? For example, do I need to
worry about all or even any of the steps that individual RNAPs go
through each time they make a transcript? Turning the question
around, can we see any imprint of those processes in the
available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other
beautifully here.
\end{itemize}}
\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As shown in Figure~1.
\mmnote{Key ideas:
\begin{itemize}
\item Exclusive repression and occupancy hypothesis. Not
necessarily steric hindrance, e.g., maybe kinked DNA, but it is
key that there is zero transcription when repressor is bound.
This is the key takeaway from the qualitative cartoon in Fig 1A.
\item Change order in Fig~1 and order of presentation of noneq
models to match my treatment of Fano factors. This doesn't do
them in order of increasing math complexity, but it does do them
in logical order: Poisson, RNAP bound/unbound states, RNAP
unbound/closed/open states, ambiguous active/inactive, bursty
Poisson. For each model, sketch the intuition for its
construction and interpretation of the states/weights/rates,
possibly calling ahead/deferring some arguments to the
constitutive Fano section. Then quote the expression for mean
mRNA (that will be derived more carefully in an appendix), then
show the algebra to compute fold-change from mean mRNA and show
how it fits into form of the master curve.
\item Emphasize that tuning any of the parameters in $\rho$ is
probably off-limits experimentally: it wrecks entire cell
physiology and confounds everything. You can write down models of
what you THINK the important steps are and aren't, and use in
vitro data for the important steps, but without any way to
experimentally perturb in vivo, you have no way to TEST whether
your modeling assumptions are correct! Also point out
supercoiling~\cite{Chong2014,Sevier2016} to emphasize that the
potential features that are necessary to include are WAY more
than just RNAP details. One possible tuning knob is maybe RNAP
binding sites, but we don't have a terrific predictive
understanding of how mutations there would influence any of the
parameters inside $\rho$, so still not obvious how to make use of
that. Brewster/Daniel2012 energy matrix is only good in one model
framework and doesn't tell you anything about how model params in
other pictures would respond.
\item Weak promoter corresponds to $\rho\approx 1$. In principle,
if repressor binding E or kinetics are known to high enough
precision, deviations from $\rho=1$ become a matter of expt. In
practice, I don't think it's at all realistic to measure
repressor params or expression levels precisely enough to detect
the expected tiny deviations from $\rho=1$. We can estimate: how
good would determination of $\Delta F_R$ in Fig~1B have to be to
get a given uncertainty in $\rho$? I think it's approx 1-to-1:
e.g., an uncertainty of 0.1 (kT) in $\Delta F_R$ converts to a
0.1 uncert in $\log\rho$, and since $\rho\approx 1$, that means
$\log(1+\Delta\rho)\approx\Delta\rho\approx 0.1$.
That's a really hard determination of $\Delta F_R$ for a very
coarse determination of $\rho$. Doubtful that inference of
kinetic rates would be any easier and probably a lot harder.Also,
in vitro RNAP data says $\rho\approx 1$ should be true, and even
if that's the wrong quantity to be talking about, Manuel's
inference of $k^+$ and $k^-$ suggests that still $\rho$ is close
to 1, and that inference is agnostic to the microscopics.
\item Master curve exists b/c repressor is on its own leaf of
graph. (But maybe don't use that language.) Arb noneq stuff can
happen on rest of promoter state space, doesn't matter. Is this
``one curve to rule them all'' a tautology? How do we probe the
theory? Can't do it at level of means, there's no discriminatory
power, need to go beyond. Test if free energy inferred from
fold-change measurements agrees with TF rates inferred from
single molec, and if those rates agree with rates inferred from
mRNA population distributions.
\end{itemize}
\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.pdf}
\caption{\textbf{
    An overview of the simple repression motif at the level of means.}
    (A) depicts the qualitative biological picture of simple repression.
    (B) shows the ``master curve'' to which all cartoons in (C) and (D) collapse.
    (C) shows two possible equilibrium pictures of simple repression, while
    (D) shows four possible coarse-graining in a nonequilibrium picture.
\label{fig1:means_cartoons}
\end{figure}

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig1.5/fig1point5.ai}
\caption{\textbf{Constitutive promoter cartoon comparison.}
The left column depicts various plausible cartoons for constitutive promoters.
We also list literature references where models have been used previously.
In model (1), transcripts are produced in a Poisson process~\cite{Jones2014}.
Model (2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}.
Model (3) is a more detailed generalization of model (2),
treating transcription initiation as a multi-step process
proceeding through closed and open complexes~\cite{Mitarai2015}.
Model (4) is somewhat analogous to (2) except with the precise nature
of active and inactive states left ambiguous~\cite{Razo-Mejia2020}.
Finally, model (5) can be viewed as a certain limit of model (4)
in which transcripts are produced in bursts,
and initiation of bursts is a Poisson process.
\mmnote{Haven't found model 5 formulated quite like this in the literature,
usually burstiness is handled like model 4, but I should probably search a bit more.}
The right column shows the Fano factor for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$
strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data
observed in~\cite{Jones2014} without introducing additional
assumptions and model complexity.
}
\label{fig:constit_cartoons}
\end{figure}

@@

In fact, we suspect \textit{any} model in which transcription
proceeds through a multistep cycle must necessarily have $\nu<1$.
The intuitive argument compares the waiting time distribution to
traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple
exponentials, and intuitively should be more peaked with a
smaller fractional width than an exponential with the same mean.\footnote{
This can be made more precise with a result
from~\cite{Stewart2007}, who showed that the convolution of
multiple gamma distributions (of which the exponential
distribution is a special case) is to a very good approximation
also gamma distributed. Using their Eq.~2 for the distribution of
the convolution, with shape parameters set to 1 to
give exponential distributions, the total waiting time
distribution has a ratio of variance to squared mean 
$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$,
where the $k_i$ are the rates of the individual steps. Clearly
this is less than 1 and therefore the total waiting time
distribution is narrower than the corresponding exponential.} A
less disperse waiting time distribution means transcription
initations are more uniformly distributed in time relative to a
Poisson process. Hence the distribution of mRNA over a population
of cells should be less variable compared to Poisson, giving
$\nu<1$.\footnote{
This last step, while intuitive, can be argued by analogy to
photon statistics where antibunching gives rise to sub-Poissonian
noise~\cite{Paul1982, Zou1990}. Although loopholes exist, we do
not expect they apply for our problem. Nevertheless we refrain
from elevating this cycle/sub-Poissonian equivalence to a ``theorem.''
\mmnote{Antibunching is the ``obvious'' analogy for me but it'd
be totally out of left field for most readers, any suggestions on
a more comprehensible reference??}
}

@@

While this seems promising, there is a major drawback for our purposes
which was already uncovered by the authors of~\cite{Razo-Mejia2020}:
the statistical inference problem is nonidentifiable,
in the sense described in Section 4.3 of~\cite{Gelman2013},
meaning it is impossible to infer the
parameters $r$ and $k^-$ from the FISH data of~\cite{Jones2014}
(as shown in Fig.~S2 of~\cite{Razo-Mejia2020}).
Rather, only the ratio $r/k^-$ could be inferred. In that work,
the problem was worked around with an informative prior on the
ratio $k^-/k^+$. That approach is unlikely to work here, as,
recall, our entire goal in modeling constitutive expression is to
use it as the basis for a yet more complicated model, when we add
on repression. But adding more complexity to a model that is
already poorly identified is a fool's errand,
so we will explore one more potential model.
